---
title: "Final_TMT_MSstats_analysis"
author: "Sam Siljee"
date: "2025-01-04"
output: html_document
---

Created: 2025/01/04
Written by: Sam Siljee
copyright (c) - 2025 Sam Siljee

# Introduction
This script was used to run the analysis on my final TMT proteomics data.

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(44)

# Set directories and annotations file name
input_dir <- paste0(getwd(), "/TMT_input_filtered/")
output_dir <- paste0(getwd(), "/TMT_output_weighted/")
run_annotations_file <- "run_annotations.tsv"
channel_annotations_file <- "channel_annotations_subgroups.tsv"
contrast_matrix_file <- "contrast_matrix_subgroups.tsv"
load_from_rda <- TRUE

# Load packages
library(MSstatsTMT) # Mass spec data statistical analysis
# library(devtools) # Install MSstatsWeightedSummary from local folder
# devtools::install_local("MSstatsWeightedSummary-devel/") # Install MSstatsWeightedSummary if needed
library(MSstatsWeightedSummary) # Used to weigh shared peptides for quantification
library(vroom) # Fast data loading
library(dplyr) # Data manipulation
library(tidyr) # Split column into multiple rows for weighted peptides input
library(stringr) # Renaming PSM column names
library(data.table) # Convert data.frame to data.table for weighted peptide input
library(splitstackshape) # Split up the protein groups
```

# Functions
```{r Functions, include=FALSE}
# Function to import the data given a data type key
import_psm_data <- function(input_dir, file_list) {
  # Filter files list to only include relevant files
  filtered_file_list <- grep(file_list, pattern = paste0("_PSMs.txt$"), value = TRUE)

  # initialise data.frame
  dat <- data.frame()

  # Loop through and import files - add index to connect with annotations
  for (i in filtered_file_list) {
    dat <- bind_rows(
      dat,
      vroom(i)
    )
  }

  # return combined data.frame
  return(dat)
}
```

# Read in the data
MS runs initially analysed in Proteome Discoverer 2.4, PSM files exported as text files for further processing here with `MSstatsTMT`.
Annotation file manually created.
The data is imported, then formatted for `MSstatsTMT` using `PDtoMSstatsTMTFormat`.
```{r data import}
if (load_from_rda) { # Load from .rda files
  load(paste0(input_dir, "PSMs.rda"))
  load(paste0(input_dir, "annotations.rda"))
} else { # Load from .csv files
  # Load PSM data data
  PSMs <- import_psm_data(input_dir, list.files(input_dir, full.names = TRUE))
  colnames(PSMs) <- colnames(PSMs) %>% str_replace_all(" ", ".")
  # Save the PSMs data for faster future input
  save(PSMs, file = paste0(input_dir, "PSMs.rda"))

  # Load annotations file
  annotations <- full_join(
    read.delim(paste0(input_dir, run_annotations_file), header = TRUE, stringsAsFactors = TRUE),
    read.delim(paste0(input_dir, channel_annotations_file), header = TRUE, stringsAsFactors = TRUE)
  )
  # Save the annotations data for faster future input
  save(annotations, file = paste0(input_dir, "annotations.rda"))
}

# Load contrast matrix file
contrast_matrix <- read.delim(paste0(input_dir, contrast_matrix_file), header = TRUE)
row.names(contrast_matrix) <- contrast_matrix[, 1]
contrast_matrix <- contrast_matrix[, -1]
```

# Format input
```{r}
# Format input using MSstatsTMT - will also deal with fractions
MSstatsTMT_input <- MSstatsTMT::PDtoMSstatsTMTFormat(
  input = PSMs,
  annotation = annotations,
  which.proteinid = "Protein.Accessions",
  useNumProteinsColumn = FALSE, # Keep shared peptides
  useUniquePeptide = FALSE, # Keep shared peptides
  rmPSM_withfewMea_withinRun = FALSE, # Default: TRUE
  rmProtein_with1Feature = FALSE,
  summaryforMultipleRows = sum,
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/PDtoMSstatsTMTFormat_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Save the formatted input
save(MSstatsTMT_input, file = paste0(output_dir, "MSstatsTMT_input.rda"))
```

# Weighted preprocessing
Method based on this vignette: https://vitek-lab.github.io/MSstatsWeightedSummary/articles/shared_workflow.html
```{r}
# First option to preprocess input
# Select required columns
required_input <- as.data.table(MSstatsTMT_input)[, .(ProteinName, PeptideSequence, Charge, PSM, Run, 
                           Channel, Intensity, Condition, BioReplicate,
                           Mixture, TechRepMixture)]

# Add clusters
pp_graph <- createPeptideProteinGraph(required_input)
required_input <- addClusterMembership(required_input, pp_graph)

# Run normalisation
required_input_1 <- normalizeSharedPeptides(required_input)

# Second option to preprocess input
# Process the input
MSstatsTMT_processed <- MSstatsTMT::proteinSummarization(
  data = MSstatsTMT_input,
  method = "msstats",
  global_norm = TRUE,
  reference_norm = TRUE,
  remove_norm_channel = TRUE,
  remove_empty_channel = TRUE,
  MBimpute = TRUE,
  maxQuantileforCensored = NULL,
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/proteinSummarization_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  ),
  msstats_log_path = paste0(
    output_dir,
    "/logs/MSstats_proteinSummarization_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Filter to only include required columns
required_input_2 <- as.data.table(mutate(MSstatsTMT_processed$FeatureLevelData, Intensity = 2 ^ log2Intensity))[, .(ProteinName, PeptideSequence, Charge, PSM, Run, 
                           Channel, Intensity, Condition, BioReplicate,
                           Mixture, TechRepMixture)]

# Run weighting
MSstatsTMT_summarised_1 <- getWeightedProteinSummary(
  required_input_1,
  norm = "Huber",
  norm_parameter = 0.1,
  max_iter = 30)

MSstatsTMT_summarised_2 <- getWeightedProteinSummary(
  required_input_2,
  norm = "Huber",
  norm_parameter = 0.1,
  max_iter = 30)

# Generate inputs for group comparisons
MSstatsTMT_processed_1 <- makeMSstatsTMTInput(MSstatsTMT_summarised_1)
MSstatsTMT_processed_2 <- makeMSstatsTMTInput(MSstatsTMT_summarised_2)
```

# Model-based comparison, and adjusted p-value
```{r model based comparison, echo = FALSE}
# Run the test
MSstatsTMT_test <- groupComparisonTMT(
  contrast.matrix = contrast_matrix,
  data = MSstatsTMT_processed,
  moderated = FALSE, # Steph: TRUE
  adj.method = "BH",
  remove_norm_channel = TRUE,
  remove_empty_channel = TRUE,
  save_fitted_models = FALSE,
  use_log_file = TRUE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/groupComparison_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Extracting the comparison results
MSstatsTMT_results <- MSstatsTMT_test$ComparisonResult

# Checking for outliers, results in which there is an infinite fold-change
outliers <- which(MSstatsTMT_results$log2FC == Inf | MSstatsTMT_results$log2FC == -Inf)
paste("There are", length(outliers), "proteins found only in one condition, where the adjusted P value is zero, and fold-change is infinite")

# save the results
save(MSstatsTMT_results, file = paste0(output_dir, "MSstatsTMT_results.rda"))
write.csv(MSstatsTMT_results, file = paste0(output_dir, "MSstatsTMTResults.csv"))
```

# Session information
Save file of session information
```{r SessionInfo}
writeLines(
  capture.output(sessionInfo()),
  paste0(
    output_dir,
    "logs/sessionInfo_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)
```
