---
title: "Final_TMT_phos_MSstats_analysis"
author: "Sam Siljee"
date: "2025-02-03"
output: html_document
---

Created: 2025/01/04
Written by: Sam Siljee
copyright (c) - 2025 Sam Siljee

# Introduction
This script was used to run the analysis on my final TMT phosphoproteomics data.

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(44)

# Set directories and annotations file name
input_dir <- paste0(getwd(), "/Phos_input/")
output_dir <- paste0(getwd(), "/Phos_output/")
run_annotations_phos_file <- "run_annotations_phos.tsv"
run_annotations_proteins_file <- "run_annotations_proteins.tsv"
channel_annotations_file <- "channel_annotations_subgroups.tsv"
contrast_matrix_file <- "contrast_matrix_subgroups.tsv"
FASTA_file <- "Human_swissprot_no_iso_20250915_0602_Cell_Culture_Contaminants_GFP_to_eGFP_edit.fasta"
prob_threshold <- 75

# Load packages
library(MSstatsPTM) # Mass spec data statisitcal analysis
library(MSstatsConvert) # Prepare input data for MSstats
library(vroom) # Fast data loading
library(dplyr) # Data manipulation
library(stringr) # String manipulation
library(tidyr) # For unnest_wider function

# Function to import the data given a data type key
import_psm_data <- function(file_list) {
  # Filter files list to only include relevant files
  filtered_file_list <- grep(file_list, pattern = paste0("_PSMs.txt$"), value = TRUE)

  # initialise data.frame
  dat <- data.frame()

  # Loop through and import files - add index to connect with annotations
  for (i in filtered_file_list) {
    dat <- bind_rows(
      dat,
      vroom(i)
    )
  }

  # return combined data.frame
  return(dat)
}

# Function to parse phosphosite probabilities
parse_phos_confidence <- function(x) {
  # Return NA if missing or no phospho data
  if (is.na(x) || !grepl(pattern = "(Phospho): ", x, fixed = TRUE)) {
    return(list(phosphosite = NA, probability = NA))
  } else {
    # Separate the modifications
    x <- unlist(str_split(x, "; "))
    
    # Split each modification on "(Phospho): "
    parts <- strsplit(x, "(Phospho): ", fixed = TRUE)
    
    # Get residue numbers and probability scores
    res   <- sapply(parts, `[`, 1)
    probs <- as.numeric(sapply(parts, `[`, 2))
    
    # Find highest probability
    best_prob <- max(probs)
    
    # Concatenate all sites matching probability
    best_res  <- paste(res[probs == best_prob], collapse = ";")
    
    return(list(phosphosite = best_res, probability = best_prob))
  }
}
```

# Read in the data
MS runs initially analysed in Proteome Discoverer 2.4, PSM files exported as text files for further processing here with `MSstatsTMT`.
Annotation file manually created.
The data is imported, then formatted for `MSstatsTMT` using `PDtoMSstatsTMTFormat`.
```{r data import}
# Load PSM data data
PSMs_phos <- import_psm_data(paste0(input_dir, grep("Phospho", list.files(input_dir), value = TRUE)))
PSMs_proteins <- import_psm_data(paste0(input_dir, grep("Phospho", list.files(input_dir), value = TRUE, invert = TRUE)))

# Replace spaces with periods - MSstats expects "R-friendly headers" from PSM output
colnames(PSMs_phos) <- colnames(PSMs_phos) %>% str_replace_all(" ", ".")
colnames(PSMs_proteins) <- colnames(PSMs_proteins) %>% str_replace_all(" ", ".")

# Parse the site probability column and filter
PSMs_phos_filtered <- PSMs_phos %>%
  rowwise() %>%
  mutate(parsed = list(parse_phos_confidence(`ptmRS:.Best.Site.Probabilities`))) %>%
  unnest_wider(parsed) %>%
  filter(probability > prob_threshold)

# Load annotations file
annotations_phos <- full_join(
  read.delim(paste0(input_dir, run_annotations_phos_file), header = TRUE, stringsAsFactors = TRUE),
  read.delim(paste0(input_dir, channel_annotations_file), header = TRUE, stringsAsFactors = TRUE)
)
annotations_proteins <- full_join(
  read.delim(paste0(input_dir, run_annotations_proteins_file), header = TRUE, stringsAsFactors = TRUE),
  read.delim(paste0(input_dir, channel_annotations_file), header = TRUE, stringsAsFactors = TRUE)
)

# Load contrast matrix file
contrast_matrix <- read.delim(paste0(input_dir, contrast_matrix_file), header = TRUE)
row.names(contrast_matrix) <- contrast_matrix[,1]
contrast_matrix <- contrast_matrix[,-1]
contrast_matrix <- as.matrix(contrast_matrix)

# Format for `MSstatsTMT` input
MSstatsPTM_input <- MSstatsPTM::PDtoMSstatsPTMFormat(
  input = PSMs_phos,
  annotation = annotations_phos,
  fasta_path = paste0(input_dir, FASTA_file),
  protein_input = PSMs_proteins,
  annotation_protein = annotations_proteins,
  labeling_type = "TMT",
  mod_id = "\\(Phospho\\)",
  use_localization_cutoff = FALSE,
  keep_all_mods = FALSE,
  use_unmod_peptides = FALSE,
  fasta_protein_name = "uniprot_iso",
  localization_cutoff = 75,
  remove_unlocalized_peptides = TRUE,
  useNumProteinsColumn = FALSE,
  useUniquePeptide = TRUE,
  summaryforMultipleRows = max,
  removeFewMeasurements = TRUE,
  removeOxidationMpeptides = FALSE,
  removeProtein_with1Peptide = FALSE,
  which_quantification = "Precursor.Area",
  which_proteinid = "Master.Protein.Accessions",
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/PDtoMSstatsPTMFormat_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

MSstatsPTM_input_filtered <- MSstatsPTM::PDtoMSstatsPTMFormat(
  input = PSMs_phos_filtered,
  annotation = annotations_phos,
  fasta_path = paste0(input_dir, FASTA_file),
  protein_input = PSMs_proteins,
  annotation_protein = annotations_proteins,
  labeling_type = "TMT",
  mod_id = "\\(Phospho\\)",
  use_localization_cutoff = FALSE,
  keep_all_mods = FALSE,
  use_unmod_peptides = FALSE,
  fasta_protein_name = "uniprot_iso",
  localization_cutoff = 75,
  remove_unlocalized_peptides = TRUE,
  useNumProteinsColumn = FALSE,
  useUniquePeptide = TRUE,
  summaryforMultipleRows = max,
  removeFewMeasurements = TRUE,
  removeOxidationMpeptides = FALSE,
  removeProtein_with1Peptide = FALSE,
  which_quantification = "Precursor.Area",
  which_proteinid = "Master.Protein.Accessions",
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/PDtoMSstatsPTMFormat_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Save the prepared data
save(MSstatsPTM_input, file = paste0(output_dir, "PTM_preprocessed.rda"))
save(MSstatsPTM_input_filtered, file = paste0(output_dir, "PTM_preprocessed_filtered.rda"))
```

# Data Processing
This chunk processes the input data including normalisation, censored cutoff, cleaning, and protein-level summarization.
```{r data processing, echo = FALSE}
# Process the input
MSstatsPTM_processed <- MSstatsPTM::dataSummarizationPTM_TMT(
  data = MSstatsPTM_input,
  method = "msstats",
  global_norm = TRUE,
  global_norm.PTM = TRUE,
  reference_norm = TRUE,
  reference_norm.PTM = TRUE,
  remove_norm_channel = TRUE,
  remove_empty_channel = TRUE,
  MBimpute = TRUE,
  MBimpute.PTM = TRUE,
  maxQuantileforCensored = NULL,
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/proteinSummarization_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Save the processed data
save(MSstatsPTM_processed, file = paste0(output_dir, "PTM_processed.rda"))

# Process the input
MSstatsPTM_processed_filtered <- MSstatsPTM::dataSummarizationPTM_TMT(
  data = MSstatsPTM_input_filtered,
  method = "msstats",
  global_norm = TRUE,
  global_norm.PTM = TRUE,
  reference_norm = TRUE,
  reference_norm.PTM = TRUE,
  remove_norm_channel = TRUE,
  remove_empty_channel = TRUE,
  MBimpute = TRUE,
  MBimpute.PTM = TRUE,
  maxQuantileforCensored = NULL,
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/proteinSummarization_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Save the processed data
save(MSstatsPTM_processed_filtered, file = paste0(output_dir, "PTM_processed_filtered.rda"))

# Repeat without normalisation for normalisation plot
# Process the input
MSstatsPTM_processed_no_normalisation <- MSstatsPTM::dataSummarizationPTM_TMT(
  data = MSstatsPTM_input,
  method = "msstats",
  global_norm = FALSE,
  global_norm.PTM = FALSE,
  reference_norm = FALSE,
  reference_norm.PTM = FALSE,
  remove_norm_channel = TRUE,
  remove_empty_channel = TRUE,
  MBimpute = TRUE,
  MBimpute.PTM = TRUE,
  maxQuantileforCensored = NULL,
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/proteinSummarization_log_no_normalisation_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Save the processed data
save(MSstatsPTM_processed_no_normalisation, file = paste0(output_dir, "PTM_processed_no_normalisation.rda"))
```

# Model-based comparison, and adjusted p-value
```{r model based comparison, echo = FALSE}
# Run the test
test_MSstatsPTM <- groupComparisonPTM(
  data = MSstatsPTM_processed,
  data.type = "TMT",
  contrast.matrix = contrast_matrix,
  moderated = FALSE, # Steph: TRUE
  adj.method = "BH",
  use_log_file = TRUE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/groupComparison_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Run the test
test_MSstatsPTM_filtered <- groupComparisonPTM(
  data = MSstatsPTM_processed_filtered,
  data.type = "TMT",
  contrast.matrix = contrast_matrix,
  moderated = FALSE, # Steph: TRUE
  adj.method = "BH",
  use_log_file = TRUE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/groupComparison_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Extracting the comparison results
MSstatsPTM_results <- test_MSstatsPTM$ADJUSTED.Model
MSstatsPTM_raw_results <- test_MSstatsPTM$PTM.Model
MSstatsPTM_results_filtered <- test_MSstatsPTM_filtered$ADJUSTED.Model
MSstatsPTM_raw_results_filtered <- test_MSstatsPTM_filtered$PTM.Model

# save the results
save(MSstatsPTM_results, file = paste0(output_dir, "MSstatsPTM_results.rda"))
write.csv(MSstatsPTM_results, file = paste0(output_dir, "MSstatsPTMResults.csv"))
save(MSstatsPTM_raw_results, file = paste0(output_dir, "MSstatsPTM_raw_results.rda"))
write.csv(MSstatsPTM_raw_results, file = paste0(output_dir, "MSstatsPTMrawResults.csv"))
save(MSstatsPTM_results_filtered, file = paste0(output_dir, "MSstatsPTM_results_filtered.rda"))
write.csv(MSstatsPTM_results_filtered, file = paste0(output_dir, "MSstatsPTMResults_filtered.csv"))
save(MSstatsPTM_raw_results_filtered, file = paste0(output_dir, "MSstatsPTM_raw_results_filtered.rda"))
write.csv(MSstatsPTM_raw_results_filtered, file = paste0(output_dir, "MSstatsPTMrawResults_filtered.csv"))
```

# Session information
Save file of session information
```{r SessionInfo}
writeLines(
  capture.output(sessionInfo()),
  paste0(
    output_dir,
    "logs/sessionInfo_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
    )
  )
```
