---
title: "Final_TMT_MSstats_analysis"
author: "Sam Siljee"
date: "2025-01-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Created: 2024/11/22 
Written by: Sam Siljee
copyright (c) - 2024 Sam Siljee

# Introduction
This script was used to run the analysis on my final LFQ proteomics data.

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set directories and annotations file name
input_dir <- paste0(getwd(), "/TMT_input/")
output_dir <- paste0(getwd(), "/TMT_output/")
annotations_file <- "quick_run_annotations.tsv"
contrast_matrix_file <- "quick_contrast_matrix.tsv"

# Load packages
library(MSstatsTMT) # Mass spec data statisitcal analysis
library(MSstatsConvert) # Prepare input data for MSstats
library(vroom) # Fast data loading
library(dplyr) # Data manipulation
library(ggplot2) # Plotting

# Get the number of available cores, reserve 20% for other tasks
available_cores <- floor(0.8 * parallel::detectCores())

# Function to import the data given a data type key
import_psm_data <- function(input_dir, file_list) {
  # Filter files list to only include relevant files
  filtered_file_list <- grep(file_list, pattern = paste0("_PSMs.txt$"), value = TRUE)

  # initialise data.frame
  dat <- data.frame()

  # Loop through and import files - add index to connect with annotations
  for (i in filtered_file_list) {
    dat <- bind_rows(
      dat,
      vroom(i)
    )
  }

  # return combined data.frame
  return(dat)
}
```

# Read in the data
MS runs initially analysed in Proteome Discoverer 2.4, PSM files exported as text files for further processing here with `MSstats`.
Annotation file manually created.
The data is imported, then formatted for `MSstats` using `PDtoMSstatsFormat`.

```{r data import}
# Load PD LFQ output from txt files
files_list <- list.files(input_dir, full.names = TRUE)

# Load PSM data data
PSMs <- import_psm_data(input_dir, list.files(input_dir, full.names = TRUE))

# Load annotations file
annotations <- read.delim(paste0(input_dir, annotations_file), header = TRUE, stringsAsFactors = TRUE)

# Load contrast matrix file
contrast_matrix <- read.delim(paste0(input_dir, contrast_matrix_file), header = TRUE)
row.names(contrast_matrix) <- contrast_matrix[,1]
contrast_matrix <- contrast_matrix[,-1]

# Rename the columns required for the `PDtoMSstatsFormat` call
PSMs <- PSMs %>% rename(
  # ProteinGroupAccessions = `Master Protein Accessions`,
  ProteinGroupAccessions = `Protein Accessions`,
  PrecursorArea = `Precursor Abundance`,
  SpectrumFile = `Spectrum File`
)

# Format for `MSstats` input
MSstats_input <- MSstatsConvert::PDtoMSstatsFormat(
  input = PSMs,
  annotation = annotations,
  useNumProteinsColumn = FALSE,
  useUniquePeptide = TRUE,
  summaryforMultipleRows = max,
  removeFewMeasurements = FALSE, # Changed from default
  removeOxidationMpeptides = FALSE,
  removeProtein_with1Peptide = FALSE,
  which.quantification = "PrecursorArea",
  which.proteinid = "ProteinGroupAccessions",
  which.sequence = "AnnotatedSequence",
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/PDtoMSstatsFormat_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Briefly check input
head(MSstats_input)

# Count the number of proteins
paste("There are", (MSstats_input$ProteinName %>% unique() %>% length()), "proteins in this dataset")

# Save the prepared data
save(MSstats_input, file = paste0(output_dir, "preprocessed.rda"))
```
