---
title: "Final_TMT_MSstats_analysis"
author: "Sam Siljee"
date: "2025-01-04"
output: html_document
---

Created: 2025/01/04
Written by: Sam Siljee
copyright (c) - 2025 Sam Siljee

# Introduction
This script was used to run the analysis on my final TMT proteomics data.

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(44)

# Set directories and annotations file name
input_dir <- paste0(getwd(), "/TMT_input/")
output_dir <- paste0(getwd(), "/TMT_output/")
run_annotations_file <- "run_annotations.tsv"
channel_annotations_file <- "channel_annotations.tsv"
contrast_matrix_file <- "contrast_matrix.tsv"

# Load packages
library(MSstatsTMT) # Mass spec data statisitcal analysis
library(MSstatsWeightedSummary) # Used to weigh shared peptides for quantification
library(vroom) # Fast data loading
library(dplyr) # Data manipulation
```

# Functions
```{r Functions, include=FALSE}
# Function to import the data given a data type key
import_psm_data <- function(input_dir, file_list) {
  # Filter files list to only include relevant files
  filtered_file_list <- grep(file_list, pattern = paste0("_PSMs.txt$"), value = TRUE)

  # initialise data.frame
  dat <- data.frame()

  # Loop through and import files - add index to connect with annotations
  for (i in filtered_file_list) {
    dat <- bind_rows(
      dat,
      vroom(i)
    )
  }

  # return combined data.frame
  return(dat)
}

# Function to split protein groups
split_protein_groups <- function(PSMs_input) {
  # Initialise dataframe
  dat <- data.frame()
  
  # Loop through rows of input and split up proteins from groups if needed
  for(i in 1:nrow(PSMs_input)){
    # Extract current row
    current_row <- PSMs_input[i,]
    
    if(grepl(current_row$`Protein Accessions`, pattern = ";")){ # If current row is a protein group, split it up
      # Make list of proteins in the group
      group_proteins <- current_row$`Protein Accessions` %>% str_split("; ") %>% unlist()
      # Loop through proteins and add a row to the data frame for each
      for(j in group_proteins) {
        dat <- rbind(dat, mutate(current_row, `Protein Accessions` = j))
      }
      
    } else { # If not a protein group just add to dataframe as-is
      dat <- rbind(dat, current_row)
    }
  }
  
  # Return the dataframe
  return(dat)
}
```

# Read in the data
MS runs initially analysed in Proteome Discoverer 2.4, PSM files exported as text files for further processing here with `MSstatsTMT`.
Annotation file manually created.
The data is imported, then formatted for `MSstatsTMT` using `PDtoMSstatsTMTFormat`.

```{r data import}
# Load PSM data data
PSMs <- import_psm_data(input_dir, list.files(input_dir, full.names = TRUE))

# Load annotations file
annotations <- full_join(
  read.delim(paste0(input_dir, run_annotations_file), header = TRUE, stringsAsFactors = TRUE),
  read.delim(paste0(input_dir, channel_annotations_file), header = TRUE, stringsAsFactors = TRUE)
)

# Load contrast matrix file
contrast_matrix <- read.delim(paste0(input_dir, contrast_matrix_file), header = TRUE)
row.names(contrast_matrix) <- contrast_matrix[,1]
contrast_matrix <- contrast_matrix[,-1]
```

# Weighted preprocessing
Preprocessing the TMT input using the shared peptide weighting method
```{r Weighted input preprocessing}
# Re-format the PSMs data to split up protein groups

curve_pp_raw = unique(PSMs[, .(PeptideSequence = `Annotated Sequence`, ProteinName = `Protein Accessions`)])
curve_pp = cSplit(curve_pp_raw, splitCols = "ProteinName", sep = "; ", direction = "long", drop = FALSE)
curve_pp[, ProteinName := stringr::str_replace_all(ProteinName, "\\-1", "")]
curve_pp_graph = createPeptideProteinGraph(curve_pp)
```

# Weighted processing
Processing the TMT input using the shared peptide weighting method
```{r Weighted input processing}

```

# Preprocessing
Data preprocisseing using the original MSstatsTMT workflow
```{r Input preprocessing}
# Format for `MSstatsTMT` input
MSstatsTMT_input <- MSstatsTMT::PDtoMSstatsTMTFormat(
  input = PSMs_weighted_summary,
  annotation = annotations,
  which.proteinid = "Master Protein Accessions", # Default: Protein Accessions
  useNumProteinsColumn = TRUE,
  useUniquePeptide = TRUE,
  rmPSM_withfewMea_withinRun = FALSE, # Default: TRUE
  rmProtein_with1Feature = FALSE,
  summaryforMultipleRows = sum,
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/PDtoMSstatsTMTFormat_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Briefly check input
head(MSstatsTMT_input)

# Count the number of proteins
paste("There are", (MSstatsTMT_input$ProteinName %>% unique() %>% length()), "proteins in this dataset")

# Save the prepared data
save(MSstatsTMT_input, file = paste0(output_dir, "TMT_preprocessed.rda"))
```

# Data Processing
This chunk processes the input data including normalisation, censored cutoff, cleaning, and protein-level summarization.
Using the original MSstatsTMT method
```{r data processing, echo = FALSE}
# Process the input
MSstatsTMT_processed <- MSstatsTMT::proteinSummarization(
  data = MSstatsTMT_input,
  method = "msstats",
  global_norm = TRUE,
  reference_norm = TRUE,
  remove_norm_channel = TRUE,
  remove_empty_channel = TRUE,
  MBimpute = TRUE,
  maxQuantileforCensored = NULL,
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/proteinSummarization_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  ),
  msstats_log_path = paste0(
    output_dir,
    "/logs/MSstats_proteinSummarization_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Save the processed data
save(MSstatsTMT_processed, file = paste0(output_dir, "TMT_processed.rda"))

# Repeat without normalisation for normalisation plot
# Process the input - without normalisation
MSstatsTMT_processed_no_normalisation <- MSstatsTMT::proteinSummarization(
  data = MSstatsTMT_input,
  method = "msstats",
  global_norm = FALSE,
  reference_norm = FALSE,
  remove_norm_channel = TRUE,
  remove_empty_channel = TRUE,
  MBimpute = TRUE,
  maxQuantileforCensored = NULL,
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/proteinSummarization_no_normalisation_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  ),
  msstats_log_path = paste0(
    output_dir,
    "/logs/MSstats_proteinSummarization_no_normalisation_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Save the processed data
save(MSstatsTMT_processed_no_normalisation, file = paste0(output_dir, "TMT_processed_no_normalisation.rda"))
```

# Data visualisation
These plots are from functions built-in to `MSstatsTMT` and are for quality control.
```{r data visualisation, echo = FALSE}
# # Produces individual plots for the different proteins, each peptide shown in a different colour
# dataProcessPlotsTMT(
#   data=MSstatsTMT_processed,
#   type='ProfilePlot',
#   width = 7,
#   height = 7,
#   address = paste0(output_dir, "MSstatsTMT_")
# )
# 
# # This produces a boxplot, of intensity by sample. Used as a basic way to check normalisation
# dataProcessPlotsTMT(
#   data=MSstatsTMT_processed,
#   type='QCPlot',
#   width = 7,
#   height = 7,
#   address = paste0(output_dir, "MSstatsTMT_")
# )
```

# Model-based comparison, and adjusted p-value

```{r model based comparison, echo = FALSE}
# Run the test
test_MSstatsTMT <- groupComparisonTMT(
  contrast.matrix = contrast_matrix,
  data = MSstatsTMT_processed,
  moderated = FALSE, # Steph: TRUE
  adj.method = "BH",
  remove_norm_channel = TRUE,
  remove_empty_channel = TRUE,
  save_fitted_models = FALSE,
  use_log_file = TRUE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/groupComparison_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Extracting the comparison results
MSstatsTMT_results <- test_MSstatsTMT$ComparisonResult

# Checking for outliers, results in which there is an infinite fold-change
outliers <- which(MSstatsTMT_results$log2FC == Inf | MSstatsTMT_results$log2FC == -Inf)
paste("There are", length(outliers), "proteins found only in one condition, where the adjusted P value is zero, and fold-change is infinite")

# save the results
save(MSstatsTMT_results, file = paste0(output_dir, "MSstatsTMT_results.rda"))
write.csv(MSstatsTMT_results, file = paste0(output_dir, "MSstatsTMTResults.csv"))
```

# Session information
Save file of session information
```{r SessionInfo}
writeLines(
  capture.output(sessionInfo()),
  paste0(
    output_dir,
    "logs/sessionInfo_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
    )
  )
```
