---
title: "Final_TMT_MSstats_analysis"
author: "Sam Siljee"
date: "2025-01-04"
output: html_document
---

Created: 2025/01/04
Written by: Sam Siljee
copyright (c) - 2025 Sam Siljee

# Introduction
This script was used to run the analysis on my final TMT proteomics data.

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(44)

# Set directories and annotations file name
input_dir <- paste0(getwd(), "/TMT_input_no_isoforms/")
output_dir <- paste0(getwd(), "/TMT_output_weighted/")
run_annotations_file <- "run_annotations.tsv"
channel_annotations_file <- "channel_annotations_subgroups.tsv"
contrast_matrix_file <- "contrast_matrix_subgroups.tsv"
load_from_rda <- TRUE

# Load packages
library(MSstatsTMT) # Mass spec data statistical analysis
# library(devtools) # Install MSstatsWeightedSummary from local folder
# devtools::install_local("MSstatsWeightedSummary-devel/") # Install MSstatsWeightedSummary if needed
library(MSstatsWeightedSummary) # Used to weigh shared peptides for quantification
library(vroom) # Fast data loading
library(dplyr) # Data manipulation
library(tidyr) # Split column into multiple rows for weighted peptides input
library(stringr) # Renaming PSM column names
library(data.table) # Convert data.frame to data.table for weighted peptide input
library(splitstackshape) # Split up the protein groups
```

# Functions
```{r Functions, include=FALSE}
# Function to import the data given a data type key
import_psm_data <- function(input_dir, file_list) {
  # Filter files list to only include relevant files
  filtered_file_list <- grep(file_list, pattern = paste0("_PSMs.txt$"), value = TRUE)

  # initialise data.frame
  dat <- data.frame()

  # Loop through and import files - add index to connect with annotations
  for (i in filtered_file_list) {
    dat <- bind_rows(
      dat,
      vroom(i)
    )
  }

  # return combined data.frame
  return(dat)
}

# Function to select highest fraction
filter_highest_fraction <- function(PSMs) {
  # Add index column and filter for selected rows
  dat <- mutate(PSMs, index = 1:nrow(PSMs))
  
  # Summarise and get list of indexes of maximum
  
  # Return PSM list filtered by top indexes
}
```

# Read in the data
MS runs initially analysed in Proteome Discoverer 2.4, PSM files exported as text files for further processing here with `MSstatsTMT`.
Annotation file manually created.
The data is imported, then formatted for `MSstatsTMT` using `PDtoMSstatsTMTFormat`.
```{r data import}
if (load_from_rda) { # Load from .rda files
  load(paste0(input_dir, "PSMs.rda"))
  load(paste0(input_dir, "annotations.rda"))
} else { # Load from .csv files
  # Load PSM data data
  PSMs <- import_psm_data(input_dir, list.files(input_dir, full.names = TRUE))
  colnames(PSMs) <- colnames(PSMs) %>% str_replace_all(" ", ".")
  # Save the PSMs data for faster future input
  save(PSMs, file = paste0(input_dir, "PSMs.rda"))

  # Load annotations file
  annotations <- full_join(
    read.delim(paste0(input_dir, run_annotations_file), header = TRUE, stringsAsFactors = TRUE),
    read.delim(paste0(input_dir, channel_annotations_file), header = TRUE, stringsAsFactors = TRUE)
  )
  # Save the annotations data for faster future input
  save(annotations, file = paste0(input_dir, "annotations.rda"))
}

# Load contrast matrix file
contrast_matrix <- read.delim(paste0(input_dir, contrast_matrix_file), header = TRUE)
row.names(contrast_matrix) <- contrast_matrix[, 1]
contrast_matrix <- contrast_matrix[, -1]
```

# Weighted preprocessing - not working
Preprocessing the TMT input using the shared peptide weighting method - based on the MWS reproduction github page:
https://github.com/mstaniak/MWS_reproduction/blob/main/05_tpp_onepot_data_preparation.R
```{r Weighted input preprocessing}
# Process using weighting for unique peptides
# Format using MSstatsTMT - will also deal with fractions
MSstatsTMT_input <- MSstatsTMT::PDtoMSstatsTMTFormat(
  input = PSMs,
  annotation = annotations,
  which.proteinid = "Protein.Accessions",
  useNumProteinsColumn = FALSE, # Keep shared peptides
  useUniquePeptide = FALSE, # Keep shared peptides
  rmPSM_withfewMea_withinRun = FALSE, # Default: TRUE
  rmProtein_with1Feature = FALSE,
  summaryforMultipleRows = sum,
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/PDtoMSstatsTMTFormat_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
) %>%
  as.data.table()

# Save for troubleshooting
save(MSstatsTMT_input, file = paste0(output_dir, "MSstats_input_1.rda"))

# Split the protein groups
MSstatsTMT_input <- cSplit(MSstatsTMT_input, sep = "; ", direction = "long", drop = FALSE, splitCols = "ProteinName")

# Save for troubleshooting
save(MSstatsTMT_input, file = paste0(output_dir, "MSstats_input_1b.rda"))

# Add clusters for the processIsoforms function
pp_graph <- createPeptideProteinGraph(MSstatsTMT_input)
MSstatsTMT_input <- addClusterMembership(MSstatsTMT_input, pp_graph)

# Save for troubleshooting
save(MSstatsTMT_input, file = paste0(output_dir, "MSstats_input_2.rda"))

# Process isoforms
MSstatsTMT_input <- processIsoforms(MSstatsTMT_input, TRUE, TRUE, FALSE)

# Save for troubleshooting
save(MSstatsTMT_input, file = paste0(output_dir, "MSstats_input_3.rda"))

# Normalise
MSstatsTMT_input <- normalizeSharedPeptides(MSstatsTMT_input)

# Save for troubleshooting
save(MSstatsTMT_input, file = paste0(output_dir, "MSstats_input_4.rda"))

# Add protein group clusters
pp_graph <- createPeptideProteinGraph(MSstatsTMT_input)
MSstatsTMT_input <- addClusterMembership(MSstatsTMT_input, pp_graph)

# Save for troubleshooting
save(MSstatsTMT_input, file = paste0(output_dir, "MSstats_input_5.rda"))

# Remove protein duplicates
MSstatsTMT_input <- unique(MSstatsTMT_input)

# Save for troubleshooting
save(MSstatsTMT_input, file = paste0(output_dir, "MSstats_input_6.rda"))

# Save the prepared data
save(MSstatsTMT_input, file = paste0(output_dir, "TMT_preprocessed_weighted.rda"))
```


# Weighted preprocessing - Staniak method
```{r}
psms <- as.data.table(PSMs)
pp_orig <- unique(psms[, .(PeptideSequence = `Annotated.Sequence`, ProteinName = `Protein.Accessions`)])
pp <- cSplit(pp_orig, sep = "; ", direction = "long", drop = FALSE, splitCols = "ProteinName")
pp[, ProteinName := stringr::str_replace_all(ProteinName, "\\-1", "")]
pp_graph <- createPeptideProteinGraph(pp)

rm(pp_orig)
gc()

# # Cluster identification ----
pp <- addClusterMembership(pp, pp_graph)
input_all_prots <- merge(psms, pp, by.x = "Annotated.Sequence", by.y = "PeptideSequence", allow.cartesian = T, all.x = T, all.y = T)
input_all_prots[, `Master.Protein.Accessions` := NULL]
input_all_prots[, `Protein.Accessions` := NULL]

rm(pp, pp_graph, psms)
gc()

# # MSstatsTMT pre-processing ----
procd <- MSstatsTMT::PDtoMSstatsTMTFormat(input_all_prots, annotations,
                                          which.proteinid = "ProteinName",
                                          useNumProteinsColumn = FALSE,
                                          useUniquePeptide = FALSE,
                                          rmPSM_withfewMea_withinRun = FALSE,
                                          rmProtein_with1Feature = FALSE,
                                          summaryforMultipleRows = max,
                                          use_log_file = FALSE,
                                          append = FALSE,
                                          verbose = TRUE
)

rm(input_all_prots)
gc()

procd <- as.data.table(procd)
# Process isoforms ---
# ## onePot
procd[, log2Intensity := log(Intensity, 2)]
graph_procd <- createPeptideProteinGraph(procd)
procd <- addClusterMembership(procd, graph_procd)
procd[, IsUnique := uniqueN(ProteinName) == 1, by = "PSM"]

# procd_merge_single_identical <- processIsoforms(procd, T, T, F)
# procd_merge_single_identical[, IsUnique := uniqueN(ProteinName) == 1, by = "PSM"]

# Optional steps:
# merging or filtering isoforms if there is not enough feature-level information
procd_iso <- processIsoforms(procd, T, T, F)

rm(procd, graph_procd)
gc()

# Normalization (feature level, median-based)
procd_iso[, Intensity := 2^log2Intensity]
procd_iso <- normalizeSharedPeptides(procd_iso)
procd_iso_graph <- createPeptideProteinGraph(procd_iso)
procd_iso <- addClusterMembership(procd_iso, procd_iso_graph)
# Cluster statistics
procd_iso <- getClusterStatistics(procd_iso, TRUE)

rm(procd_iso_graph)
gc()

MSstatsTMT_input <- unique(procd_iso)

# Save the prepared data
save(MSstatsTMT_input, file = paste0(output_dir, "TMT_preprocessed_weighted.rda"))
```

# Weighted processing
Processing the TMT input using the shared peptide weighting method
```{r Weighted input processing}
# Get protein summaries
MSstats_summarised <- getWeightedProteinSummary(
  MSstatsTMT_input,
  norm = "p_norm",
  norm_parameter = 1,
  weights_mode = "contributions",
  tolerance = 0.1,
  max_iter = 10,
  initial_summary = "unique",
  weights_penalty = FALSE,
  weights_penalty_param = 0.1,
  save_weights_history = FALSE,
  save_convergence_history = FALSE
)

MSstatsTMT_processed <- makeMSstatsTMTInput(MSstats_summarised)

# Save the processed data
save(MSstatsTMT_processed, file = paste0(output_dir, "TMT_processed_weighted.rda"))
```

# Preprocessing
Data preprocisseing using the original MSstatsTMT workflow
```{r Input preprocessing}
# # Format for `MSstatsTMT` input
# MSstatsTMT_input <- MSstatsTMT::PDtoMSstatsTMTFormat(
#   input = PSMs,
#   annotation = annotations,
#   which.proteinid = "Master Protein Accessions", # Default: Protein Accessions
#   useNumProteinsColumn = TRUE,
#   useUniquePeptide = TRUE,
#   rmPSM_withfewMea_withinRun = FALSE, # Default: TRUE
#   rmProtein_with1Feature = FALSE,
#   summaryforMultipleRows = sum,
#   use_log_file = TRUE,
#   append = FALSE,
#   verbose = TRUE,
#   log_file_path = paste0(
#     output_dir,
#     "/logs/PDtoMSstatsTMTFormat_log_",
#     format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
#     ".txt"
#   )
# )
#
# # Briefly check input
# head(MSstatsTMT_input)
#
# # Count the number of proteins
# paste("There are", (MSstatsTMT_input$ProteinName %>% unique() %>% length()), "proteins in this dataset")
#
# # Save the prepared data
# save(MSstatsTMT_input, file = paste0(output_dir, "TMT_preprocessed.rda"))
```

# Data Processing
This chunk processes the input data including normalisation, censored cutoff, cleaning, and protein-level summarization.
Using the original MSstatsTMT method
```{r data processing, echo = FALSE}
# # Process the input
# MSstatsTMT_processed <- MSstatsTMT::proteinSummarization(
#   data = MSstatsTMT_input,
#   method = "msstats",
#   global_norm = TRUE,
#   reference_norm = TRUE,
#   remove_norm_channel = TRUE,
#   remove_empty_channel = TRUE,
#   MBimpute = TRUE,
#   maxQuantileforCensored = NULL,
#   use_log_file = TRUE,
#   append = FALSE,
#   verbose = TRUE,
#   log_file_path = paste0(
#     output_dir,
#     "/logs/proteinSummarization_log_",
#     format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
#     ".txt"
#   ),
#   msstats_log_path = paste0(
#     output_dir,
#     "/logs/MSstats_proteinSummarization_log_",
#     format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
#     ".txt"
#   )
# )
#
# # Save the processed data
# save(MSstatsTMT_processed, file = paste0(output_dir, "TMT_processed.rda"))
#
# # Repeat without normalisation for normalisation plot
# # Process the input - without normalisation
# MSstatsTMT_processed_no_normalisation <- MSstatsTMT::proteinSummarization(
#   data = MSstatsTMT_input,
#   method = "msstats",
#   global_norm = FALSE,
#   reference_norm = FALSE,
#   remove_norm_channel = TRUE,
#   remove_empty_channel = TRUE,
#   MBimpute = TRUE,
#   maxQuantileforCensored = NULL,
#   use_log_file = TRUE,
#   append = FALSE,
#   verbose = TRUE,
#   log_file_path = paste0(
#     output_dir,
#     "/logs/proteinSummarization_no_normalisation_log_",
#     format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
#     ".txt"
#   ),
#   msstats_log_path = paste0(
#     output_dir,
#     "/logs/MSstats_proteinSummarization_no_normalisation_log_",
#     format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
#     ".txt"
#   )
# )
#
# # Save the processed data
# save(MSstatsTMT_processed_no_normalisation, file = paste0(output_dir, "TMT_processed_no_normalisation.rda"))
```

# Data visualisation
These plots are from functions built-in to `MSstatsTMT` and are for quality control.
```{r data visualisation, echo = FALSE}
# # Produces individual plots for the different proteins, each peptide shown in a different colour
# dataProcessPlotsTMT(
#   data=MSstatsTMT_processed,
#   type='ProfilePlot',
#   width = 7,
#   height = 7,
#   address = paste0(output_dir, "MSstatsTMT_")
# )
#
# # This produces a boxplot, of intensity by sample. Used as a basic way to check normalisation
# dataProcessPlotsTMT(
#   data=MSstatsTMT_processed,
#   type='QCPlot',
#   width = 7,
#   height = 7,
#   address = paste0(output_dir, "MSstatsTMT_")
# )
```

# Model-based comparison, and adjusted p-value

```{r model based comparison, echo = FALSE}
# Run the test
MSstatsTMT_test <- groupComparisonTMT(
  contrast.matrix = contrast_matrix,
  data = MSstatsTMT_processed,
  moderated = FALSE, # Steph: TRUE
  adj.method = "BH",
  remove_norm_channel = TRUE,
  remove_empty_channel = TRUE,
  save_fitted_models = FALSE,
  use_log_file = TRUE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/groupComparison_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Extracting the comparison results
MSstatsTMT_results <- MSstatsTMT_test$ComparisonResult

# Checking for outliers, results in which there is an infinite fold-change
outliers <- which(MSstatsTMT_results$log2FC == Inf | MSstatsTMT_results$log2FC == -Inf)
paste("There are", length(outliers), "proteins found only in one condition, where the adjusted P value is zero, and fold-change is infinite")

# save the results
save(MSstatsTMT_results, file = paste0(output_dir, "MSstatsTMT_results.rda"))
write.csv(MSstatsTMT_results, file = paste0(output_dir, "MSstatsTMTResults.csv"))
```

# Session information
Save file of session information
```{r SessionInfo}
writeLines(
  capture.output(sessionInfo()),
  paste0(
    output_dir,
    "logs/sessionInfo_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)
```
