---
title: "Final_TMT_phos_MSstats_analysis"
author: "Sam Siljee"
date: "2025-02-03"
output: html_document
---

Created: 2025/01/04
Written by: Sam Siljee
copyright (c) - 2025 Sam Siljee

# Introduction
This script was used to run the analysis on my final TMT phosphoproteomics data.

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(44)

# Set directories and annotations file name
input_dir <- paste0(getwd(), "/TMT_phos_input/")
output_dir <- paste0(getwd(), "/TMT_phos_output/")
run_annotations_phos_file <- "run_annotations_phos.tsv"
run_annotations_proteins_file <- "run_annotations_proteins.tsv"
channel_annotations_file <- "channel_annotations.tsv"
contrast_matrix_file <- "contrast_matrix.tsv"
FASTA_file <- "Human_Swissprot_w_contaminants_eGFP_20250115.fasta"

# Load packages
library(MSstatsPTM) # Mass spec data statisitcal analysis
library(MSstatsConvert) # Prepare input data for MSstats
library(vroom) # Fast data loading
library(dplyr) # Data manipulation

# Function to import the data given a data type key
import_psm_data <- function(file_list) {
  # Filter files list to only include relevant files
  filtered_file_list <- grep(file_list, pattern = paste0("_PSMs.txt$"), value = TRUE)

  # initialise data.frame
  dat <- data.frame()

  # Loop through and import files - add index to connect with annotations
  for (i in filtered_file_list) {
    dat <- bind_rows(
      dat,
      vroom(i)
    )
  }

  # return combined data.frame
  return(dat)
}
```

# Read in the data
MS runs initially analysed in Proteome Discoverer 2.4, PSM files exported as text files for further processing here with `MSstatsTMT`.
Annotation file manually created.
The data is imported, then formatted for `MSstatsTMT` using `PDtoMSstatsTMTFormat`.

```{r data import}
# Load PSM data data
PSMs_phos <- import_psm_data(paste0(input_dir, grep("Phospho", list.files(input_dir), value = TRUE)))
PSMs_proteins <- import_psm_data(paste0(input_dir, grep("Phospho", list.files(input_dir), value = TRUE, invert = TRUE)))

# Replace spaces with periods - MSstats expects "R-friendly headers" from PSM output
colnames(PSMs_phos) <- colnames(PSMs_phos) %>% str_replace_all(" ", ".")
colnames(PSMs_proteins) <- colnames(PSMs_proteins) %>% str_replace_all(" ", ".")

# Load annotations file
annotations_phos <- full_join(
  read.delim(paste0(input_dir, run_annotations_phos_file), header = TRUE, stringsAsFactors = TRUE),
  read.delim(paste0(input_dir, channel_annotations_file), header = TRUE, stringsAsFactors = TRUE)
)
annotations_proteins <- full_join(
  read.delim(paste0(input_dir, run_annotations_proteins_file), header = TRUE, stringsAsFactors = TRUE),
  read.delim(paste0(input_dir, channel_annotations_file), header = TRUE, stringsAsFactors = TRUE)
)

# Load contrast matrix file
contrast_matrix <- read.delim(paste0(input_dir, contrast_matrix_file), header = TRUE)
row.names(contrast_matrix) <- contrast_matrix[,1]
contrast_matrix <- contrast_matrix[,-1]

# Format for `MSstatsTMT` input
MSstatsPTM_input <- MSstatsPTM::PDtoMSstatsPTMFormat(
  input = PSMs_phos,
  annotation = annotations_phos,
  fasta_path = paste0(input_dir, FASTA_file),
  protein_input = PSMs_proteins,
  annotation_protein = annotations_proteins,
  labeling_type = "TMT",
  mod_id = "\\(Phospho\\)",
  use_localization_cutoff = FALSE,
  keep_all_mods = FALSE,
  use_unmod_peptides = FALSE,
  fasta_protein_name = "uniprot_iso",
  localization_cutoff = 75,
  remove_unlocalized_peptides = TRUE,
  useNumProteinsColumn = FALSE,
  useUniquePeptide = TRUE,
  summaryforMultipleRows = max,
  removeFewMeasurements = TRUE,
  removeOxidationMpeptides = FALSE,
  removeProtein_with1Peptide = FALSE,
  which_quantification = "Precursor.Area",
  which_proteinid = "Master.Protein.Accessions",
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/PDtoMSstatsPTMFormat_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Briefly check input
head(MSstatsPTM_input)

# Save the prepared data
save(MSstatsPTM_input, file = paste0(output_dir, "PTM_preprocessed.rda"))
```

# Data Processing
This chunk processes the input data including normalisation, censored cutoff, cleaning, and protein-level summarization.
```{r data processing, echo = FALSE}
# Process the input
MSstatsPTM_processed <- MSstatsPTM::dataSummarizationPTM_TMT(
  data = MSstatsPTM_input,
  method = "msstats",
  global_norm = TRUE,
  global_norm.PTM = TRUE,
  reference_norm = TRUE,
  reference_norm.PTM = TRUE,
  remove_norm_channel = TRUE,
  remove_empty_channel = TRUE,
  MBimpute = TRUE,
  MBimpute.PTM = TRUE,
  maxQuantileforCensored = NULL,
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/proteinSummarization_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Save the processed data
save(MSstatsPTM_processed, file = paste0(output_dir, "PTM_processed.rda"))

# Repeat without normalisation for normalisation plot
# Process the input
MSstatsPTM_processed_no_normalisation <- MSstatsPTM::dataSummarizationPTM_TMT(
  data = MSstatsPTM_input,
  method = "msstats",
  global_norm = FALSE,
  global_norm.PTM = FALSE,
  reference_norm = FALSE,
  reference_norm.PTM = FALSE,
  remove_norm_channel = TRUE,
  remove_empty_channel = TRUE,
  MBimpute = TRUE,
  MBimpute.PTM = TRUE,
  maxQuantileforCensored = NULL,
  use_log_file = TRUE,
  append = FALSE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/proteinSummarization_log_no_normalisation_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Save the processed data
save(MSstatsPTM_processed_no_normalisation, file = paste0(output_dir, "PTM_processed_no_normalisation.rda"))
```

# Data visualisation
These plots are from functions built-in to `MSstatsPTM` and are for quality control.
```{r data visualisation, echo = FALSE}
# Produces individual plots for the different proteins, each peptide shown in a different colour
dataProcessPlotsPTM(
  data=MSstatsPTM_processed,
  type='ProfilePlot',
  width = 7,
  height = 7,
  address = paste0(output_dir, "MSstatsPTM_")
)

# This produces a boxplot, of intensity by sample. Used as a basic way to check normalisation
dataProcessPlotsPTM(
  data=MSstatsPTM_processed,
  type='QCPlot',
  width = 7,
  height = 7,
  address = paste0(output_dir, "MSstatsPTM_")
)
```

# Model-based comparison, and adjusted p-value
```{r model based comparison, echo = FALSE}
# Run the test
test_MSstatsPTM <- groupComparisonPTM(
  data = MSstatsPTM_processed,
  data.type = "TMT",
  contrast.matrix = contrast_matrix,
  moderated = FALSE, # Steph: TRUE
  adj.method = "BH",
  use_log_file = TRUE,
  verbose = TRUE,
  log_file_path = paste0(
    output_dir,
    "/logs/groupComparison_log_",
    format(Sys.time(), "%Y_%m_%d_%H_%M_%S"),
    ".txt"
  )
)

# Extracting the comparison results
MSstatsPTM_results <- test_MSstatsPTM$ADJUSTED.Model
MSstatsPTM_raw_results <- test_MSstatsPTM$PTM.Model

# Add column for up/downregulated or non-significant
MSstatsPTM_results$Dif <- ifelse(MSstatsPTM_results$log2FC > 1 & MSstatsPTM_results$adj.pvalue < 0.05,
  "Upregulated",
  ifelse(MSstatsPTM_results$log2FC < -1 & MSstatsPTM_results$adj.pvalue < 0.05,
    "Downregulated",
    "Not significant"
  )
)

# save the results
save(MSstatsPTM_results, file = paste0(output_dir, "MSstatsPTM_results.rda"))
write.csv(MSstatsPTM_results, file = paste0(output_dir, "MSstatsPTMResults.csv"))
save(MSstatsPTM_raw_results, file = paste0(output_dir, "MSstatsPTM_raw_results.rda"))
write.csv(MSstatsPTM_raw_results, file = paste0(output_dir, "MSstatsPTMrawResults.csv"))
```
