---
title: "Untitled"
author: "Sam Siljee"
date: '2022-11-18'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MSstats)
library(tidyverse)
```

# Data import
This workflow is to import data at the PSM level from Proteome Discoverer 2.4 (PD).
After running the analysis in PD, select File > Export > To Text (tab delimited)...
Save the file and place it in the input 

```{r data import, echo = FALSE}
# read main data
raw_IT <- read.csv("~/Coding/MSstats-tutorial/data/OT_vs_IT/Pilot_22091_1ul_50cm_IT_PSMs.txt", sep = "\t", stringsAsFactors = TRUE)
raw_OT <- read.csv("~/Coding/MSstats-tutorial/data/OT_vs_IT/Pilot_22091_1ul_50cm_OT_PSMs.txt", sep = "\t", stringsAsFactors = TRUE)
raw <- rbind(raw_IT, raw_OT)

# read annotations
annot <- read.csv("~/Coding/MSstats-tutorial/data/OT_vs_IT/OT_vs_IT_annotation.csv", sep = ",", stringsAsFactors = TRUE)

# format for `MSstats`
# currently this is not working. I suspect there's an issue with the column names of the PSMs export from PD not matching the expected column names in the `PDtoMSstatsFormat` call. See more here; https://groups.google.com/g/msstats/c/FR4-69Z_Neo
input <- PDtoMSstatsFormat(raw, 
                           annotation=annot,
                           removeProtein_with1Peptide=FALSE)

head(input)

## count the number of proteins
input$ProteinName %>% unique %>% length
```

# Data Processing
including Normalization, decide censored cutoff, protein-level summarization
Argument deleted, to be compatible with newer version of `MSstats`; 'cutoffCensored = "minFeature"'
See discussion; https://groups.google.com/g/msstats/c/WLZQz1-QdiA/m/3yjqu4TBAAAJ

```{r data processing, echo = FALSE}
processed.quant <- dataProcess(input,
                               normalization = 'equalizeMedians',
                               summaryMethod="TMP",
                               censoredInt="NA",
                               MBimpute=TRUE,
                               maxQuantileforCensored=0.999)

save(processed.quant, file='data/Proteome_Discoverer_example/output/processed.quant.rda')
```

# Data visualisation

```{r data visualisation, echo = FALSE}
dataProcessPlots(processed.quant, type="QCplot", 
                 ylimDown=0, 
                 which.Protein = 'allonly',
                 width=7, height=7,  
                 address="data/Proteome_Discoverer_example/output/ControlMixture_DDA_ProteomeDiscoverer_")

dataProcessPlots(processed.quant, type="Profileplot", 
                 ylimDown=0, 
                 originalPlot = TRUE,
                 summaryPlot = TRUE,
                 width=7, height=7,  
                 address="data/Proteome_Discoverer_example/output/ControlMixture_DDA_ProteomeDiscoverer_")

dataProcessPlots(processed.quant, type="Conditionplot", 
                 ylimDown=0, 
                 width=7, height=7,  
                 address="data/Proteome_Discoverer_example/output/ControlMixture_DDA_ProteomeDiscoverer_")
```

## Model-based comparison, and adjusted p-value

```{r model based comparison, echo = FALSE}
# generate compariosn matrix
comparison1<-matrix(c(1,-1,0,0,0),nrow=1)
comparison2<-matrix(c(1,0,-1,0,0),nrow=1)
comparison3<-matrix(c(1,0,0,-1,0),nrow=1)
comparison4<-matrix(c(1,0,0,0,-1),nrow=1)
comparison5<-matrix(c(0,1,-1,0,0),nrow=1)
comparison6<-matrix(c(0,1,0,-1,0),nrow=1)
comparison7<-matrix(c(0,1,0,0,-1),nrow=1)
comparison8<-matrix(c(0,0,1,-1,0),nrow=1)
comparison9<-matrix(c(0,0,1,0,-1),nrow=1)
comparison10<-matrix(c(0,0,0,1,-1),nrow=1)
comparison<-rbind(comparison1,comparison2, comparison3, comparison4, comparison5, 
                  comparison6, comparison7, comparison8, comparison9, comparison10)
row.names(comparison)<-c("Condition1-Condition2", "Condition1-Condition3", "Condition1-Condition4", "Condition1-Condition5", "Condition2-Condition3", 
                         "Condition2-Condition4", "Condition2-Condition5", "Condition3-Condition4", "Condition3-Condition5", "Condition4-Condition5")
colnames(comparison) <- paste("Condition", 1:5, sep = "")

test.MSstats <- groupComparison(contrast.matrix=comparison, data=processed.quant)
test.MSstats <- test.MSstats$ComparisonResult

# save the results
save(test.MSstats, file='data/Proteome_Discoverer_example/output/test.MSstats.rda')
write.csv(test.MSstats, file='data/Proteome_Discoverer_example/output/ControlMixture_DDA_ProteomeDiscoverer_testResult_byMSstats.csv')
```

# Comparison plots

```{r comparison plots, echo = FALSE}
groupComparisonPlots(data=test.MSstats, type="VolcanoPlot",
                     width=6, height=6,
                     address="data/Proteome_Discoverer_example/output/ControlMixture_DDA_ProteomeDiscoverer_")

groupComparisonPlots(data=test.MSstats, type="ComparisonPlot",
                     width=6, height=6,
                     address="data/Proteome_Discoverer_example/output/ControlMixture_DDA_ProteomeDiscoverer_")

```
